The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-18 paper/ms.tex
Add new chervin midplane plot to paper

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on the true action value. 
A midplane error of $\sim6\,\pc$ for a thin disk orbit or
$\sim50\,\pc$ for a thick disk orbit is all that is necessary to
induce a $10\,\%$ error in the vertical action $J_z$.
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the
induced action error from either a midplane offset or a Galactocentric radius
offset. We make the epicyclic approximation, which assumes that the motion in
the $z$ and $R$ components of the orbit are decoupled and follow a simple
harmonic oscillator. This approximation should be an excellent description of
thin disk-like orbits and a good description of thick disk-like orbits. We
also make the assumption of a perfectly flat circular velocity curve with
velocity $v_c$.

Under this approximation, we can write down the vertical and radial components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
z(t) &= A_z \sin{(\kappa t + \alpha)}\\
R(t) &= A_R \sin{(\nu t + \beta)}\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. We can write down the actions in this approximation as
\citep[Section~3.5.3b][]{2008gady.book.....B},
\beq\label{eq:actions_epi}
\begin{split}
L_z &= R_g v_c \\
J_r &= \kappa A_R^2 \\
J_z &= \nu A_z^2\text{,}
\end{split}
\eeq
where $R_g$ is the guiding radius of the orbit.

An induced error in $z$ of $\Delta z$ will cause $\Delta A_z \sim \Delta z$.
Similarly, an induced error in $R$ will cause $\Delta A_R \sim \Delta R$. If
we assume that $\kappa$ and $\nu$ are constant (\Gus{justify?}), then we have
that
\beq\label{eq:actions_epi_error}
\begin{split}
\Delta L_z = \Delta R v_c &\implies \frac{\Delta L_z}{L_z} = \frac{\Delta R}{R} \\
\Delta J_r = 2 \kappa \Delta A_R &\implies \frac{\Delta J_r}{J_r} = \frac{2\Delta R}{A_R} \\
\Delta J_z = 2 \nu \Delta A_z &\implies \frac{\Delta J_z}{J_z} = \frac{2\Delta z}{A_z}\text{.}
\end{split}
\eeq
Therefore, errors in each action are linearly proportional to the errors in
either the midplane or cylindrical radius.

There is a slight complication in the formula for computing $\Delta J_r/J_r$ and $\Delta L_z/L_z$.
In a moment we will consider the effect of an inaccuracy in the distance from
the Sun to the Galactic center on actions. However, this inaccuracy induces an
error in $x$, or $\Delta x$. This is not the same as an inaccuracy in $\Delta
R$, but they are related by the following argument. An error $\Delta x$ will
induce an erroneous radius $R_{\text{err}}$ related by the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_r$ and $L_z$ are related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\begin{split}
\frac{\Delta J_r}{J_r} &= \frac{4 \Delta x}{\pi A_R} \\
\frac{\Delta L_z}{L_z} &= \frac{2 \Delta x}{\pi R}\text{.}
\end{split}
\eeq

We will check these formulae against numerical tests next.

@@

\begin{deluxetable*}{cccccccc}
\tablecaption{Description and names of the three orbits considered in this
work. For the last two columns: $z_{\text{max}}$ the maximum height of the
orbit while $A_R$ is the magnitude of the radial excursions of the orbit,
computed as one half the maximum radius minus the minimum
radius.\label{tab:orbits}}
\tablehead{\colhead{name} & \colhead{initial position} & \colhead{initial
velocity} & \colhead{$J_r$} &
\colhead{$L_z$} & \colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{$A_R$}\\ \colhead{ } &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{km}/\mathrm{s}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{kpc}$}}
\startdata 
thin-disk & $(8, 0, 0)$ & $(0, -190, 10)$ & 40.3 & -1520 & 0.69 & 0.12 & 1.29 \\
thick-disk & $(8, 0, 0)$ & $(0, -190, 50)$ & 32.5 & -1520 & 23.0 & 0.85 & 1.19 \\ 
halo & $(8, 0, 0)$ & $(0, -190, 190)$ & 32.8 & -1520 & 529.1 & 6.16 & 2.34
\enddata
\end{deluxetable*}

@@

\begin{figure}
\plotone{fig/schmactions_Jz_hist.pdf}
\caption{A histogram of the observed values in $J_z$ for the thick-disk orbit
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account. We also plot one half
the 95\uth{} percentile minus the 5\uth{} percentile of each distribution as a
horizontal arrow anchored on the true $J_z$ value. We call this $\Delta J_z$
and will use it (along with the similarly defined $\Delta J_r$ and $\Delta
L_z$) to empirically describe the error distribution. We see that $\Delta J_z$
roughly corresponds to the distance from the true $J_z$ value to one of the
modes of the error distribution.}
\label{fig:Jz_hist}
\end{figure}

@@

Empirically describing the error distribution shown in
Figure~\ref{fig:Jz_hist} is difficult. Of course, Gaussian summary statistics
are not adequate. We therefore elect to measure this error by computing one half the
95\uth{} percentile minus the 5\uth{} percentile of the distribution of
observed action values. We refer to this quantity as $\Delta J_i$ for each
action value. We plot this quantity in Figure~\ref{fig:Jz_hist} as a horizontal arrow anchored on the true action value. Because of the bimodality of the error distribution, this
quantity roughly measures the distance from the true action value to the peak of one of the modes.
Furthermore, this bimodality also implies that $\Delta J_i$ is not very
sensitive to the exact percentiles used. This summary statistic does not
reflect the bias induced when the midplane error is $\gtrsim z_{\text{max}}$.

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report one half the 95\uth{}
minus 5\uth{} percentile divided by the true action value ($\Delta J_i/J_i$) for the three
different fiducial orbits in Table \ref{tab:orbits}. The thick-disk
(\thickcolor) orbit is the one from Figure~\ref{fig:one_orbit_wrong_ref}, but
we also now consider the effect on the action determined for the thin-disk
(\thincolor) and halo (\halocolor) orbits.

The top row of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread induced
in each action for an offset in the $z$-component (i.e. the Galactic
midplane). In the {\em bottom row} we consider offsets in the $x$ component
(i.e. the Solar radius). The {\em left}, {\em middle}, and {\em right} columns
show the fractional spread in the values computed for $J_z$, $L_z$, and $J_r$,
respectively. We compute the fractional spread $\Delta J_i/J_i$ as one half the 95\uth{}
minus 5\uth{} percentile of the action computed using the offset coordinate
system over the course of the first $\Gyr$, divided by the true action
(computed in the correct coordinate system).

@@

\begin{figure*}
\plotone{fig/schmactions_many_orbits.pdf}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error in the
measured action ($\Delta J_i$) from coordinate system errors for the thin,
thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the text for the justification in using this to measure the magnitude of the induced error. The {\em left}, {\em center},
and {\em right} panels show the result for $J_r$, $L_z$, and $J_z$,
respectively. The {\em upper} panels consider an offset in $z$ and the {\em
lower} panels consider an offset in $x$ (equivalently, an offset in the Solar
radius).}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\plotone{fig/schmactions_many_orbits_Jz_fun.pdf}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. We show this for a $z$ offset of $10, 50, 100\,\pc$
(red, cyan, green respectively). As before, the error ($\Delta J_z$) is one half the
95\uth{} minus 5\uth{} percentile of the distribution of observed $J_z$ values
over the course of the orbit. There are large errors for thin-disk like orbits
($J_z \sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. }
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\begin{figure*}
\plotone{fig/midplane_fit_chervinsim.pdf}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Guessing here, need to
figure out what the panels actually show:}) The left two panels show the
midplane as a function of azimuth before the first encounter. As later
encounters occur, we can see that stronger midplane variations result. In
fact, one can see that an $m=2$ mode is excited in the third panel --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction).}
\label{fig:midplane_chervin}
\end{figure*}

>>>>

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on the true action value. 
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is all that is necessary to
induce a $100\,\%$ error in the vertical action $J_z$.
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@

@@

\begin{deluxetable*}{ccccccc}
\tablecaption{Description and names of the three orbits considered in this
work.\label{tab:orbits}}
\tablehead{\colhead{name} & \colhead{initial position} & \colhead{initial
velocity} & \colhead{$J_r$} &
\colhead{$L_z$} & \colhead{$J_z$} & \colhead{$z_{\text{max}}$}\\ \colhead{ } &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{km}/\mathrm{s}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc}$}}
\startdata 
thin-disk & $(8, 0, 0)$ & $(0, -190, 10)$ & 40.3 & -1520 & 0.69 & 0.12 \\
thick-disk & $(8, 0, 0)$ & $(0, -190, 50)$ & 32.5 & -1520 & 23.0 & 0.85 \\ 
halo & $(8, 0, 0)$ & $(0, -190, 190)$ & 32.8 & -1520 & 529.1 & 6.16
\enddata
\end{deluxetable*}

@@

\begin{figure}
\plotone{fig/schmactions_Jz_hist.pdf}
\caption{A histogram of the observed values in $J_z$ for the thick-disk orbit
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account.}
\label{fig:Jz_hist}
\end{figure}

@@

Empirically describing the error distribution shown in
Figure~\ref{fig:Jz_hist} is difficult. Of course, Gaussian summary statistics
are not adequate. We therefore elect to measure this error by computing the
95\uth{} percentile minus the 5\uth{} percentile of the distribution of
observed action values. We refer to this quantity as $\Delta J_i$ for each
action value. Because of the bimodality of the error distribution, this
quantity roughly measures the distance between the peaks of the two modes.
Furthermore, this bimodality also implies that $\Delta J_i$ is not very
sensitive to the exact percentiles used. This summary statistic does not
reflect the bias induced when the midplane error is $\gtrsim z_{\text{max}}$.

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report the 95\uth{}
minus 5\uth{} percentile divided by the true action value for the three
different fiducial orbits in Table \ref{tab:orbits}. The thick-disk
(\thickcolor) orbit is the one from Figure~\ref{fig:one_orbit_wrong_ref}, but
we also now consider the effect on the action determined for the thin-disk
(\thincolor) and halo (\halocolor) orbits.

The top row of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread induced
in each action for an offset in the $z$-component (i.e. the Galactic
midplane). In the {\em bottom row} we consider offsets in the $x$ component
(i.e. the Solar radius). The {\em left}, {\em middle}, and {\em right} columns
show the fractional spread in the values computed for $J_z$, $L_z$, and $J_r$,
respectively. We compute the fractional spread $\Delta J_i/J_i$ as the 95\uth{}
minus 5\uth{} percentile of the action computed using the offset coordinate
system over the course of the first $\Gyr$, divided by the true action
(computed in the correct coordinate system).

@@

\begin{figure*}
\plotone{fig/schmactions_many_orbits.pdf}
\caption{We report the 95\uth{} minus 5\uth{} percentile of the error in the
measured action ($\Delta J_i$) from coordinate system errors for the thin,
thick, and halo orbits (Table~\ref{tab:orbits}). The {\em left}, {\em center},
and {\em right} panels show the result for $J_r$, $L_z$, and $J_z$,
respectively. The {\em upper} panels consider an offset in $z$ and the {\em
lower} panels consider an offset in $x$ (equivalently, an offset in the Solar
radius).}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\plotone{fig/schmactions_many_orbits_Jz_fun.pdf}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. We show this for a $z$ offset of $10, 50, 100\,\pc$
(red, cyan, green respectively). As before, the error ($\Delta J_z$) is the
95\uth{} minus 5\uth{} percentile of the distribution of observed $J_z$ values
over the course of the orbit. There are large errors for thin-disk like orbits
($J_z \sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. }
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=5.5in]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Guessing here, need to
figure out what the panels actually show:}) The left two panels show the
midplane as a function of azimuth before the first encounter. As later
encounters occur, we can see that stronger midplane variations result. In
fact, one can see that an $m=2$ mode is excited in the third panel --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction).}
\label{fig:midplane_chervin}
\end{figure*}
